Short-term neuronal and synaptic plasticity act in synergy for deviance detection in spiking networks

### 2023-03-17 paper-1/elife.tex
Added elife version of the manuscript

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                                                 %%
%% Please do not use \input{...} to include other tex files.       %%
%% Submit your LaTeX manuscript as one .tex document.              %%
%%                                                                 %%
%% All additional figures and files should be attached             %%
%% separately and not embedded in the \TeX\ document itself.       %%
%%                                                                 %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\RequirePackage{tikz}

\documentclass[pdflatex,referee,iicol,sn-basic]{sn-jnl}

%%%% Standard Packages
\usepackage{xcolor,hyperref}
%%%%

% Make Orcid icon
\definecolor{lime}{HTML}{A6CE39}
\DeclareRobustCommand{\orcidicon}{%
	\begin{tikzpicture}
	\draw[lime, fill=lime] (0,0) 
	circle [radius=0.16] 
	node[white] {{\fontfamily{qag}\selectfont \tiny ID}};
	\draw[white, fill=white] (-0.0625,0.095) 
	circle [radius=0.007];
	\end{tikzpicture}
	\hspace{-2mm}
}

\foreach \x in {A, ..., Z}{%
	\expandafter\xdef\csname orcid\x\endcsname{\noexpand\href{https://orcid.org/\csname orcidauthor\x\endcsname}{\noexpand\orcidicon}}
}

% Define the ORCID iD command for each author separately.
\newcommand{\orcidauthorA}{0000-0002-3420-4576}
\newcommand{\orcidauthorB}{0000-0001-9402-0851}

% Shorter italic dev/msc/std
\newcommand{\dev}{\textit{dev}}
\newcommand{\msc}{\textit{msc}}
\newcommand{\std}{\textit{std}}

% Common notation
\renewcommand{\R}[3][]{{}^{#1}_{}\boldsymbol R^{#2}_{#3}}
\renewcommand{\T}[3][]{{}^{#1}_{}\boldsymbol T^{#2}_{#3}}
\newcommand{\D}[3][]{{}^{#1}_{}\boldsymbol D^{#2}_{#3}}
\newcommand{\mean}[1]{\overline{#1}}
\newcommand{\Rhat}[3][]{{}^{#1}_{}\widehat{\boldsymbol R}^{#2}_{#3}}

% Crossreferences
\newcommand{\FIG}[1]{Figure \ref{fig:#1}}
\newcommand{\EQ}[1]{Equation \ref{eq:#1}}

\jyear{2021}%

\raggedbottom
%%\unnumbered% uncomment this for unnumbered level heads

@@

\title{Short-term neuronal and synaptic plasticity act in synergy for deviance detection in spiking networks}

\author[1]{\fnm{Felix Benjamin} \sur{Kern} \orcidA{}}\email{kernfel@gmail.com}

\author*[1]{\fnm{Zenas C.} \sur{Chao} \orcidB{}}\email{zenas.c.chao@gmail.com}

\affil[1]{\orgdiv{International Research Center for Neurointelligence (WPI-IRCN)}, \orgname{The University of Tokyo}, \orgaddress{\city{Tokyo}, \country{Japan}}}

\abstract{
    Sensory areas of cortex respond more strongly to infrequent stimuli when these violate previously established regularities, a phenomenon known as deviance detection (DD). Previous modeling work has mainly attempted to explain DD on the basis of synaptic plasticity. However, a large fraction of cortical neurons also exhibit firing rate adaptation, an underexplored potential mechanism. Here, we investigate DD in a spiking neuronal network model with two types of short-term plasticity, fast synaptic short-term depression (STD) and slower threshold adaptation (TA). We probe the model with an oddball stimulation paradigm and assess DD by evaluating the network responses. We find that TA is sufficient to elicit DD. It achieves this by suppressing neurons near the stimulation site that respond earliest to the frequently presented standard stimulus (local suppression), which diminishes the response and promotes the recovery (global suppression) of the wider network. Further, we find a synergy effect between STD and TA, where they interact with each other to achieve greater DD than the sum of their individual effects. We show that this synergy is caused by the local suppression added by STD, which inhibits the global response to the frequently presented stimulus, further reducing global suppression by TA and making the network more responsive to the deviant stimulus. We conclude that highly predictable information can be encoded in strong local suppression, which allows greater global recovery and subsequent heightened sensitivity for DD.
}

\keywords{deviance detection, adaptation, short-term plasticity, neuronal plasticity, synaptic plasticity}

@@

@@

\begin{figure*}%
    \centering
    \includegraphics{fig1.pdf}
    \caption{}
    \label{fig:1}
\end{figure*}
\textbf{Fig. 1} Model and paradigm.
\textbf{a} Membrane and synaptic dynamics of a neuron innervated by an excitatory and an inhibitory neuron firing pre-determined spikes, illustrated above the plots. The excitatory connection has weight $w = 3$ for demonstration purposes only; a single connection with $w = 1$ is normally unable to evoke postsynaptic firing. Top: Inputs to the (post-)synaptic conductances $g_e$ and $g_i$ (vertical bars), and the STD depression variable $x_{Exc}$ of the excitatory presynaptic neuron (dashed line). Note that excitatory inputs are equal to $U x_{Exc} w$ as per \EQ{gsyn}. Middle: Excitatory and inhibitory synaptic conductances ($g_e$ and $g_i$, respectively). Bottom: Membrane voltage, firing threshold baseline (gray) and adaptive threshold, demonstrating TA in the excitatory postsynaptic neuron.
\textbf{b} Sample network layout, showing the spatial location of excitatory and inhibitory neurons, and all outgoing synaptic connections from two neurons of each type.
\textbf{c} Distance from stimulation site 1 in terms of minimum number of excitatory synapses. Stimulation is provided to the 10 neurons nearest to the center of each of 5 stimulation sites (large circles), which are regularly spaced on a circle with a radius of 2.5 mm.
\textbf{d} Illustration of the stimulation paradigm. Each marker represents a stimulation, with the stimulus identity indicated by color. The target stimulus (A) is represented as solid red, and non-target stimuli are represented with faded colors. Three randomized sequences were presented, containing 80\% A and 20\% B (\std{}), 20\% A and 80\% B (\dev{}), and 20\% of each of 5 stimuli, including A and B (\msc{}). Note that, while the onset asynchrony of 500 ms is represented faithfully, the horizontal extent of the stimulation markers has no meaning; stimuli were presented instantaneously rather than over an extended period

\FIG{1}a shows the internal and presynaptic dynamics of an excitatory neuron in a contrived example for the purpose of illustration. The model was implemented in Brian2 \citep{Stimberg2019-tc}, accelerated with Brian2GeNN \citep{Stimberg2020-go}, and simulated with an integration time step of 1~ms. Synaptic delay was not modeled explicitly, but spikes were delivered, and voltages reset, in the time step following spike emission. We simulated neurons without any stochasticity in either their inputs or their parameters, other than network structure (described below), in order to ensure that our results were driven purely by the experimental paradigm, rather than any model-internal sources of noise.

800 excitatory and 200 inhibitory neurons were placed randomly in a two-dimensional circular space with a radius of 4~mm, and formed synapses of weight $w = 1$ with 50 randomly selected postsynaptic partners within a range of 2~mm (excitatory) or 1~mm (inhibitory), reflecting the notion that excitatory neurons project over longer distances, while inhibitory neurons mainly connect to their local neighborhood. Connectivity and weights remained fixed throughout. Stimulation sites were evenly distributed 2.5~mm from the center of the dish, and stimulation delivered a one-time increase in $g_e$ to the 10 closest neurons, sufficient to trigger 2-3 spikes each. \FIG{1}b shows a representative example of the resulting network structure. In \FIG{1}c, we show the distance from the neurons stimulated at site 1 to each neuron in this network, in terms of the minimum number of excitatory synapses that need to be traversed to reach the target, in order to give an intuition for how activity may spread. Networks were pseudo-randomly generated according to the above scheme and screened for minimal stimulus response: Networks where fewer than 500 neurons responded to stimulation at any of the five sites after full recovery were discarded. In total, 30 networks were used for the experiments described below.

@@

To investigate deviance detection, we used a classical oddball paradigm as illustrated in \FIG{1}d, presenting a total of 500 stimuli at regular intervals of 500~ms. Target stimuli (labeled ``A'' or ``target'' hereafter) and distractor stimuli (labeled B to E) were presented in three randomized sequences: The ``standard'' sequence (\std{}), consisting of 400 presentations of A and 100 presentations of B; the ``deviant'' sequence (\dev{}) with 100 A and 400 B, and the many-standards control sequence (\msc{}) with 100 presentations of each of the stimuli A through E. In this way, A is presented as the predominant and therefore expected stimulus (\std{}), as an infrequent violation of an expectation (of B, \dev{}), and, to control for pure adaptation effects, as an equally infrequent stimulus with no strong expectation of any particular input (\msc{}).

@@

\begin{figure*}%
    \centering
    \includegraphics{fig2.pdf}
    \caption{}
    \label{fig:2}
\end{figure*}
\textbf{Fig. 2} Deviance detection in our model.
\textbf{a} Sample network target response intensity by trial type. Each horizontal line of the plot summarizes the network response in one target trial in terms of spikes per integration time step. Target trials are taken from the \std{} (400), \msc{} (100), and \dev{} (100) sequences, and sorted by sequence time (earliest trials are at the bottom of each section).
\textbf{b} Response summary, showing average network-wide spike counts in target trials in each sequence. The underlying data are the same as in panel a. Note how the \dev{} response is very large, followed by an intermediate \msc{} response and a low \std{} response.
\textbf{c} Mean response sizes $\mean{\R[dev]{A}{}}$ and $\mean{\R[std]{A}{}}$, normalized by $\mean{\R[msc]{A}{}}$, across networks and stimuli (n = 120). In this and later boxplots, the median is indicated with an orange line, notches indicate the 95\% confidence interval for the median calculated by bootstrap with 10000 iterations, boxes indicate the inter-quartile range (IQR), whiskers extend to 1.5 times the IQR or to the data extrema, whichever is less, and fliers indicate individual data points beyond the whisker end points.
\textbf{d} Deviance detection indices (DDI, see Equation {eq:ddi}) across networks and stimuli

To show that our model is capable of deviance detection, we first show the responses of the sample network in all sequences (\dev{}, \msc{}, and \std{}) in \FIG{2}a. Here, each row corresponds to one trial, aligned to stimulus input at t = 0, with the number of spikes in each millisecond bin indicated by color. All responses share an early activity pattern, which derives from the stimulated neurons and their immediate postsynaptic targets, but differences soon start to emerge. The \dev{} sequence responses, shown at the top of the plot, exhibit a clear peak in activity around 15 ms after stimulation. The \std{} sequence responses show a later, more diffuse activity peak around 20 ms, and the \msc{} sequence responses lack any obvious activity cluster. We sum these patterns up in \FIG{2}b, where we show the mean network activity, in terms of spikes per millisecond bin, across trials of each type. Clearly, the target response in the \dev{} sequence is greater than either the \msc{} or the \std{} response.

We could readily identify similar patterns emerging across most networks and stimulus assignments. To show that deviance detection is robust to network and stimulus identity, we calculated the average spike count per neuron per target trial for each network and stimulus assignment, and compared these across sequences. \FIG{2}c shows a summary of the average \dev{} and \std{} response sizes, normalized by the response in the \msc{} sequence. The data clearly confirmed that \dev{} responses were larger than \msc{} (T = 6.38e+03, p = 2.75e-13; one-sided Wilcoxon signed-rank test, n = 120), and \std{} responses smaller (T = 687, p = 6.42e-15). Accordingly, the DDI (\FIG{2}d) was positive in most networks (T = 5.97e+03, p = 4.75e-10, median 0.114). This shows that our model was well behaved and suitable for an investigation of deviance detection.

\begin{figure*}%
    \centering
    \includegraphics{fig3.pdf}
    \caption{}
    \label{fig:3}
\end{figure*}
\textbf{Fig. 3} Deviance detection under model ablation.
\textbf{a} Sample network responses to target stimulation, with the deviance detection index in each instance noted in the plot titles. Lines show the average (trial mean +- SEM, left axis) network-wide spike counts in target trials in each sequence. Boxes summarize the spike count totals per trial (right axis), the means of which form the basis for index calculations (see panel c). Note the different axis scaling on the top row (without TA) and the bottom row (with TA).
\textbf{b} Deviance detection indices across networks and stimuli. Asterisks indicate data greater than 0, established with a Wilcoxon signed-rank test at a significance level 0.05. See main text for detailed statistics

Having confirmed the presence of deviance detection in our model, we turned to an ablation approach to try to identify how the two short-term plasticity mechanisms in the model, STD and TA, enabled the networks to learn relevant regularities. \FIG{3}a shows the responses to target stimulation of the sample network under ablation. With no plasticity, the responses are indistinguishable between sequences, yielding a DDI of 0, and indicating that stimulation did not have any long-lasting effects in membrane voltage or synaptic conductance. With STD only, the response in the \dev{} sequence developed earlier and very slightly larger, yielding a DDI of 0.03. With TA only, the response became noticeably smaller, but the sequences became readily distinguishable, with high \dev{}, intermediate \msc{}, and low \std{} responses, yielding a DDI of 0.19. Finally, in the full model, the \dev{} response stood out more clearly, leaving much smaller \msc{} and \std{} responses and yielding a DDI of 0.4.

Finally, across networks and stimuli, the DDI (\FIG{3}c) was exactly 0 as expected in the ablation with no plasticity, not significantly different from 0 with STD only (T = 3.2e+03, p = 0.262, median -2.1e-04; two-sided Wilcoxon, n = 120), but clearly positive for TA only (T = 5.85e+03, p = 3.1e-09, median 0.077; one-sided Wilcoxon, n = 120) and the full model (T = 5.97e+03, p = 4.75e-10, median 0.114). To our surprise, the DDI of the full model was significantly greater than that of either plasticity mechanism alone (STD: T = 6.1e+03, p = 4.95e-11; TA: T = 4.87e+03, p = 0.000604), and greater too than the sum of the DDI values across both STD only and TA only models (T = 5.12e+03, p = 4.52e-05). This suggests that TA and STD act not as independent mechanisms in this paradigm, but rather interact constructively to enhance the network's ability to encode input regularities.

@@

To understand why the same stimulus A evoked a higher response in \dev{} than in \msc{} with TA only, we first compared the average responses in the two sequences, sorting neurons by their response onset to permit a high-resolution view of the stimulus response as it propagates through the network. As shown in \FIG{4}a (left and middle), we found that the response developed in a qualitatively similar fashion in both sequences. The contrast (right) revealed a clear increase in \dev{} spikes, particularly in the later parts of the response, as shown by the predominance of orange areas, which indicate higher \dev{} activity. Since the stimulus was identical in both sequences, and TA was the only plasticity mechanism in play, these differences had to be due to (1) different firing thresholds, and/or (2) different synaptic currents as a consequence of different prior activity within the trial.

\begin{figure*}%
    \centering
    \includegraphics{fig4.pdf}
    \caption{}
    \label{fig:4}
\end{figure*}
\textbf{Fig. 4} Responses to \dev{} target trials are larger due to lower initial $\theta$.
\textbf{a} Post-stimulus spike histogram across target trials in the sample network, showing trial time along the horizontal axis, and neurons along the vertical, sorted by the time of the first recorded spike across all trials and sequences. Left and middle column, target trials in the \dev{} sequence and \msc{} sequence, respectively; right column, contrast between these two.
\textbf{b} Target trial average of the threshold adaptation voltage $\theta$, with neuron order and columns as in panel a.
\textbf{c} Target trial average of the membrane potential $V_m$, with neuron order and columns as in panel a. The first 10 ms of the neurons under direct stimulation are masked out to avoid saturation of the colormap.
\textbf{d} Relationship across excitatory neurons of the sample network between the contrast (dev - msc) in $\theta$ at the start of target trials ($\Delta \T{A}{i}$) and the contrast in the average number of spikes fired in target trials ($\Delta \R{A}{i}$), and associated linear regression.
\textbf{e} $\Delta \T{A}{exc}$ across networks and stimuli, plotted against the Pearson correlation coefficients $\rho$ of the relationship shown in the previous panel ($\Delta \T{A}{i}$ vs $\Delta \R{A}{i}$ across excitatory neurons). Each marker corresponds to one network and target stimulus. Red markers indicate a significant negative correlation (one-sided Wald test, significance level 0.05). Inset histograms show the distribution of the corresponding values, with orange portions indicating the 95\% confidence interval of the median, calculated by bootstrap with 10000 iterations.
\textbf{f} In the sample network, contribution of the contrast in $\theta$ and $V_m$ to bins with increased firing in \dev{} (i.e., positive/red bins in the spike probability contrast, panel a), weighted by that same contrast.
\textbf{g} Contribution of $\theta$ relative to the sum of the absolute values of the weighted contributions in panel f.
\textbf{h} Relative contribution of $\theta$ across networks and stimuli, showing median (solid line) and inter-quartile range (shaded area)

We quantified the former with the strength of adaptation $\theta$ (\EQ{TA}), i.e., the amount by which the firing threshold was raised due to TA. As shown in \FIG{4}b, there was a clear and widespread lowering of thresholds at the start of target \dev{} trials relative to \msc{}, visible both in the raw data (left and middle panels) and in the contrast (right panel, early blue portions). On the other hand, once the bulk of activity had occurred, many neurons showed higher thresholds in \dev{} (right panel, later orange portions), since they spiked more often. Finally, note that inhibitory neurons, which were modeled without TA, naturally showed no difference in $\theta$, leading to numerous horizontal breaks in the plot.

To assess the difference in synaptic currents in a way that allowed direct comparison to $\theta$, we refer to the membrane voltage $V_m$, which is largely independent from and complementary to the firing threshold, and driven mostly by presynaptic inputs, resets after spikes notwithstanding. As the immediate cause of neuronal spiking, the overall pattern of $V_m$ (\FIG{4}c) largely reflected that seen in spiking activity. The contrast (right panel) shows not only a notable relative increase in membrane voltages in \dev{} around the typical response time (note the orange band running along the response front), but also an increased hyperpolarization starting immediately after due to post-spike resets.

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%% ELIFE ARTICLE TEMPLATE
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%% PREAMBLE 
\documentclass[9pt,lineno]{elife}
% Use the onehalfspacing option for 1.5 line spacing
% Use the doublespacing option for 2.0 line spacing
% Please note that these options may affect formatting.
% Additionally, the use of the \newcommand function should be limited.

\usepackage{lipsum} % Required to insert dummy text
\usepackage[version=4]{mhchem}
\usepackage{siunitx}
\DeclareSIUnit\Molar{M}

% FK: Noted, but these \newcommands are really very useful here:
% Shorter italic dev/msc/std
\newcommand{\dev}{\textit{dev}}
\newcommand{\msc}{\textit{msc}}
\newcommand{\std}{\textit{std}}

% Common notation
\newcommand{\R}[3][]{{}^{#1}_{}\boldsymbol R^{#2}_{#3}}
\newcommand{\T}[3][]{{}^{#1}_{}\boldsymbol T^{#2}_{#3}}
\newcommand{\D}[3][]{{}^{#1}_{}\boldsymbol D^{#2}_{#3}}
\newcommand{\mean}[1]{\overline{#1}}
\newcommand{\Rhat}[3][]{{}^{#1}_{}\widehat{\boldsymbol R}^{#2}_{#3}}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%% ARTICLE SETUP
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\title{Short-term neuronal and synaptic plasticity act in synergy for deviance detection in spiking networks}

\author[1]{Felix Benjamin Kern}
\author[1*]{Zenas C. Chao}
\affil[1]{International Research Center for Neurointelligence (WPI-IRCN), The University of Tokyo}

\corr{zenas.c.chao@gmail.com}{ZCC}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%% ARTICLE START
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

@@

\begin{abstract}
Sensory areas of cortex respond more strongly to infrequent stimuli when these violate previously established regularities, a phenomenon known as deviance detection (DD). Previous modeling work has mainly attempted to explain DD on the basis of synaptic plasticity. However, a large fraction of cortical neurons also exhibit firing rate adaptation, an underexplored potential mechanism. Here, we investigate DD in a spiking neuronal network model with two types of short-term plasticity, fast synaptic short-term depression (STD) and slower threshold adaptation (TA). We probe the model with an oddball stimulation paradigm and assess DD by evaluating the network responses. We find that TA is sufficient to elicit DD. It achieves this by suppressing neurons near the stimulation site that respond earliest to the frequently presented standard stimulus (local suppression), which diminishes the response and promotes the recovery (global suppression) of the wider network. Further, we find a synergy effect between STD and TA, where they interact with each other to achieve greater DD than the sum of their individual effects. We show that this synergy is caused by the local suppression added by STD, which inhibits the global response to the frequently presented stimulus, further reducing global suppression by TA and making the network more responsive to the deviant stimulus. We conclude that highly predictable information can be encoded in strong local suppression, which allows greater global recovery and subsequent heightened sensitivity for DD.
\end{abstract}

@@

\begin{figure}
    \includegraphics[width=\linewidth]{fig1}
    \caption{%
        Model and paradigm.
        \textbf{A} Membrane and synaptic dynamics of a neuron innervated by an excitatory and an inhibitory neuron firing pre-determined spikes, illustrated above the plots. The excitatory connection has weight $w = 3$ for demonstration purposes only; a single connection with $w = 1$ is normally unable to evoke postsynaptic firing. Top: Inputs to the (post-)synaptic conductances $g_e$ and $g_i$ (vertical bars), and the STD depression variable $x_{Exc}$ of the excitatory presynaptic neuron (dashed line). Note that excitatory inputs are equal to $U x_{Exc} w$ as per \EQ{gsyn}. Middle: Excitatory and inhibitory synaptic conductances ($g_e$ and $g_i$, respectively). Bottom: Membrane voltage, firing threshold baseline (gray) and adaptive threshold, demonstrating TA in the excitatory postsynaptic neuron.
        \textbf{B} Sample network layout, showing the spatial location of excitatory and inhibitory neurons, and all outgoing synaptic connections from two neurons of each type.
        \textbf{C} Distance from stimulation site 1 in terms of minimum number of excitatory synapses. Stimulation is provided to the 10 neurons nearest to the center of each of 5 stimulation sites (large circles), which are regularly spaced on a circle with a radius of 2.5 mm.
        \textbf{D} Illustration of the stimulation paradigm. Each marker represents a stimulation, with the stimulus identity indicated by color. The target stimulus (A) is represented as solid red, and non-target stimuli are represented with faded colors. Three randomized sequences were presented, containing 80\% A and 20\% B (\std{}), 20\% A and 80\% B (\dev{}), and 20\% of each of 5 stimuli, including A and B (\msc{}). Note that, while the onset asynchrony of 500 ms is represented faithfully, the horizontal extent of the stimulation markers has no meaning; stimuli were presented instantaneously rather than over an extended period.
    }
    \label{fig:1}
\end{figure}

\FIG{1}A shows the internal and presynaptic dynamics of an excitatory neuron in a contrived example for the purpose of illustration. The model was implemented in Brian2 \citep{Stimberg2019-tc}, accelerated with Brian2GeNN \citep{Stimberg2020-go}, and simulated with an integration time step of 1~ms. Synaptic delay was not modeled explicitly, but spikes were delivered, and voltages reset, in the time step following spike emission. We simulated neurons without any stochasticity in either their inputs or their parameters, other than network structure (described below), in order to ensure that our results were driven purely by the experimental paradigm, rather than any model-internal sources of noise.

800 excitatory and 200 inhibitory neurons were placed randomly in a two-dimensional circular space with a radius of 4~mm, and formed synapses of weight $w = 1$ with 50 randomly selected postsynaptic partners within a range of 2~mm (excitatory) or 1~mm (inhibitory), reflecting the notion that excitatory neurons project over longer distances, while inhibitory neurons mainly connect to their local neighborhood. Connectivity and weights remained fixed throughout. Stimulation sites were evenly distributed 2.5~mm from the center of the dish, and stimulation delivered a one-time increase in $g_e$ to the 10 closest neurons, sufficient to trigger 2-3 spikes each. \FIG{1}B shows a representative example of the resulting network structure. In \FIG{1}C, we show the distance from the neurons stimulated at site 1 to each neuron in this network, in terms of the minimum number of excitatory synapses that need to be traversed to reach the target, in order to give an intuition for how activity may spread. Networks were pseudo-randomly generated according to the above scheme and screened for minimal stimulus response: Networks where fewer than 500 neurons responded to stimulation at any of the five sites after full recovery were discarded. In total, 30 networks were used for the experiments described below.

@@

To investigate deviance detection, we used a classical oddball paradigm as illustrated in \FIG{1}D, presenting a total of 500 stimuli at regular intervals of 500~ms. Target stimuli (labeled ``A'' or ``target'' hereafter) and distractor stimuli (labeled B to E) were presented in three randomized sequences: The ``standard'' sequence (\std{}), consisting of 400 presentations of A and 100 presentations of B; the ``deviant'' sequence (\dev{}) with 100 A and 400 B, and the many-standards control sequence (\msc{}) with 100 presentations of each of the stimuli A through E. In this way, A is presented as the predominant and therefore expected stimulus (\std{}), as an infrequent violation of an expectation (of B, \dev{}), and, to control for pure adaptation effects, as an equally infrequent stimulus with no strong expectation of any particular input (\msc{}).

@@

\begin{figure}
    \includegraphics[width=\linewidth]{fig2}
    \caption{%
        Deviance detection in our model.
        \textbf{A} Sample network target response intensity by trial type. Each horizontal line of the plot summarizes the network response in one target trial in terms of spikes per integration time step. Target trials are taken from the \std{} (400), \msc{} (100), and \dev{} (100) sequences, and sorted by sequence time (earliest trials are at the bottom of each section).
        \textbf{B} Response summary, showing average network-wide spike counts in target trials in each sequence. The underlying data are the same as in panel A. Note how the \dev{} response is very large, followed by an intermediate \msc{} response and a low \std{} response.
        \textbf{C} Mean response sizes $\mean{\R[dev]{A}{}}$ and $\mean{\R[std]{A}{}}$, normalized by $\mean{\R[msc]{A}{}}$, across networks and stimuli (n = 120). In this and later boxplots, the median is indicated with an orange line, notches indicate the 95\% confidence interval for the median calculated by bootstrap with 10000 iterations, boxes indicate the inter-quartile range (IQR), whiskers extend to 1.5 times the IQR or to the data extrema, whichever is less, and fliers indicate individual data points beyond the whisker end points.
        \textbf{D} Deviance detection indices (DDI, see Equation {eq:ddi}) across networks and stimuli.
    }
    \label{fig:2}
\end{figure}

To show that our model is capable of deviance detection, we first show the responses of the sample network in all sequences (\dev{}, \msc{}, and \std{}) in \FIG{2}A. Here, each row corresponds to one trial, aligned to stimulus input at t = 0, with the number of spikes in each millisecond bin indicated by color. All responses share an early activity pattern, which derives from the stimulated neurons and their immediate postsynaptic targets, but differences soon start to emerge. The \dev{} sequence responses, shown at the top of the plot, exhibit a clear peak in activity around 15 ms after stimulation. The \std{} sequence responses show a later, more diffuse activity peak around 20 ms, and the \msc{} sequence responses lack any obvious activity cluster. We sum these patterns up in \FIG{2}B, where we show the mean network activity, in terms of spikes per millisecond bin, across trials of each type. Clearly, the target response in the \dev{} sequence is greater than either the \msc{} or the \std{} response.

We could readily identify similar patterns emerging across most networks and stimulus assignments. To show that deviance detection is robust to network and stimulus identity, we calculated the average spike count per neuron per target trial for each network and stimulus assignment, and compared these across sequences. \FIG{2}C shows a summary of the average \dev{} and \std{} response sizes, normalized by the response in the \msc{} sequence. The data clearly confirmed that \dev{} responses were larger than \msc{} (T = 6.38e+03, p = 2.75e-13; one-sided Wilcoxon signed-rank test, n = 120), and \std{} responses smaller (T = 687, p = 6.42e-15). Accordingly, the DDI (\FIG{2}D) was positive in most networks (T = 5.97e+03, p = 4.75e-10, median 0.114). This shows that our model was well behaved and suitable for an investigation of deviance detection.

\begin{figure}
    \includegraphics[width=\linewidth]{fig3}
    \caption{%
        Deviance detection under model ablation.
        \textbf{A} Sample network responses to target stimulation, with the deviance detection index in each instance noted in the plot titles. Lines show the average (trial mean +- SEM, left axis) network-wide spike counts in target trials in each sequence. Boxes summarize the spike count totals per trial (right axis), the means of which form the basis for index calculations (see panel C). Note the different axis scaling on the top row (without TA) and the bottom row (with TA).
        \textbf{B} Deviance detection indices across networks and stimuli. Asterisks indicate data greater than 0, established with a Wilcoxon signed-rank test at a significance level 0.05. See main text for detailed statistics.
    }
    \label{fig:3}
\end{figure}

Having confirmed the presence of deviance detection in our model, we turned to an ablation approach to try to identify how the two short-term plasticity mechanisms in the model, STD and TA, enabled the networks to learn relevant regularities. \FIG{3}A shows the responses to target stimulation of the sample network under ablation. With no plasticity, the responses are indistinguishable between sequences, yielding a DDI of 0, and indicating that stimulation did not have any long-lasting effects in membrane voltage or synaptic conductance. With STD only, the response in the \dev{} sequence developed earlier and very slightly larger, yielding a DDI of 0.03. With TA only, the response became noticeably smaller, but the sequences became readily distinguishable, with high \dev{}, intermediate \msc{}, and low \std{} responses, yielding a DDI of 0.19. Finally, in the full model, the \dev{} response stood out more clearly, leaving much smaller \msc{} and \std{} responses and yielding a DDI of 0.4.

Finally, across networks and stimuli, the DDI (\FIG{3}C) was exactly 0 as expected in the ablation with no plasticity, not significantly different from 0 with STD only (T = 3.2e+03, p = 0.262, median -2.1e-04; two-sided Wilcoxon, n = 120), but clearly positive for TA only (T = 5.85e+03, p = 3.1e-09, median 0.077; one-sided Wilcoxon, n = 120) and the full model (T = 5.97e+03, p = 4.75e-10, median 0.114). To our surprise, the DDI of the full model was significantly greater than that of either plasticity mechanism alone (STD: T = 6.1e+03, p = 4.95e-11; TA: T = 4.87e+03, p = 0.000604), and greater too than the sum of the DDI values across both STD only and TA only models (T = 5.12e+03, p = 4.52e-05). This suggests that TA and STD act not as independent mechanisms in this paradigm, but rather interact constructively to enhance the network's ability to encode input regularities.

@@

To understand why the same stimulus A evoked a higher response in \dev{} than in \msc{} with TA only, we first compared the average responses in the two sequences, sorting neurons by their response onset to permit a high-resolution view of the stimulus response as it propagates through the network. As shown in \FIG{4}A (left and middle), we found that the response developed in a qualitatively similar fashion in both sequences. The contrast (right) revealed a clear increase in \dev{} spikes, particularly in the later parts of the response, as shown by the predominance of orange areas, which indicate higher \dev{} activity. Since the stimulus was identical in both sequences, and TA was the only plasticity mechanism in play, these differences had to be due to (1) different firing thresholds, and/or (2) different synaptic currents as a consequence of different prior activity within the trial.

\begin{figure}
    \includegraphics[width=\linewidth]{fig4}
    \caption{%
        Responses to \dev{} target trials are larger due to lower initial $\theta$.
        \textbf{A} Post-stimulus spike histogram across target trials in the sample network, showing trial time along the horizontal axis, and neurons along the vertical, sorted by the time of the first recorded spike across all trials and sequences. Left and middle column, target trials in the \dev{} sequence and \msc{} sequence, respectively; right column, contrast between these two.
        \textbf{B} Target trial average of the threshold adaptation voltage $\theta$, with neuron order and columns as in panel A.
        \textbf{C} Target trial average of the membrane potential $V_m$, with neuron order and columns as in panel A. The first 10 ms of the neurons under direct stimulation are masked out to avoid saturation of the colormap.
        \textbf{D} Relationship across excitatory neurons of the sample network between the contrast (dev - msc) in $\theta$ at the start of target trials ($\Delta \T{A}{i}$) and the contrast in the average number of spikes fired in target trials ($\Delta \R{A}{i}$), and associated linear regression.
        \textbf{E} $\Delta \T{A}{exc}$ across networks and stimuli, plotted against the Pearson correlation coefficients $\rho$ of the relationship shown in the previous panel ($\Delta \T{A}{i}$ vs $\Delta \R{A}{i}$ across excitatory neurons). Each marker corresponds to one network and target stimulus. Red markers indicate a significant negative correlation (one-sided Wald test, significance level 0.05). Inset histograms show the distribution of the corresponding values, with orange portions indicating the 95\% confidence interval of the median, calculated by bootstrap with 10000 iterations.
        \textbf{F} In the sample network, contribution of the contrast in $\theta$ and $V_m$ to bins with increased firing in \dev{} (i.e., positive/red bins in the spike probability contrast, panel A), weighted by that same contrast.
        \textbf{G} Contribution of $\theta$ relative to the sum of the absolute values of the weighted contributions in panel F.
        \textbf{H} Relative contribution of $\theta$ across networks and stimuli, showing median (solid line) and inter-quartile range (shaded area).
    }
    \label{fig:4}
    % 
    \figsupp[The equivalent figure in the full model with both TA and STD.]{%
        The equivalent of \FIG{4} in the full model with both TA and STD.
        \textbf{A} Post-stimulus spike histogram across target trials in the sample network, showing trial time along the horizontal axis, and neurons along the vertical, sorted by the time of the first recorded spike across all trials and sequences. Left and middle column, target trials in the \dev{} sequence and \msc{} sequence, respectively; right column, contrast between these two.
        \textbf{B} Target trial average of the threshold adaptation voltage $\theta$, with neuron order and columns as in panel A.
        \textbf{C} Target trial average of the membrane potential $V_m$, with neuron order and columns as in panel A. The first 10 ms of the neurons under direct stimulation are masked out to avoid saturation of the colormap.
        \textbf{D} Relationship across excitatory neurons of the sample network between the contrast (dev - msc) in $\theta$ at the start of target trials ($\Delta \T{A}{i}$) and the contrast in the average number of spikes fired in target trials ($\Delta \R{A}{i}$), and associated linear regression.
        \textbf{E} $\Delta \T{A}{exc}$ across networks and stimuli, plotted against the Pearson correlation coefficients $\rho$ of the relationship shown in the previous panel ($\Delta \T{A}{i}$ vs $\Delta \R{A}{i}$ across excitatory neurons). Each marker corresponds to one network and target stimulus. Red markers indicate a significant negative correlation (one-sided Wald test, significance level 0.05). Inset histograms show the distribution of the corresponding values, with orange portions indicating the 95\% confidence interval of the median, calculated by bootstrap with 10000 iterations.
        \textbf{F} In the sample network, contribution of the contrast in $\theta$ and $V_m$ to bins with increased firing in \dev{} (i.e., positive/red bins in the spike probability contrast, panel A), weighted by that same contrast.
        \textbf{G} Contribution of $\theta$ relative to the sum of the absolute values of the weighted contributions in panel F.
        \textbf{H} Relative contribution of $\theta$ across networks and stimuli, showing median (solid line) and inter-quartile range (shaded area).
    }
    {\includegraphics[width=\linewidth]{fig4-withSTD}}\label{figsupp:withSTD}
\end{figure}
